---
title: "Analysis of mutations of single cell RNASeq data"
author: "Feng, Feng"
date: "`r Sys.Date()`"
header-includes:
    - \usepackage{setspace}\doublespacing
output: 
    pdf_document:
        fig_caption: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  tidy=FALSE,
  engine='R'
)
```
# 1 Aim

- Analysis the mutation analysis on the single cell RNASeq date.
- Visualize and run statistical analysis on the data of Chromosome 17 and 22.
- Also run analysis on TACI locus.

# 2 Data input

Two pieces of input are need:

- the original unprocessed single cell RNASeq data. We need
them to get the total number of sequences and total number
of cells, because we need them to normalize the 
mutation frequecy obtained below.

- Mutation counts data obtained by the SComatic software. 
The counts data were summerized and put into a spreadsheet.


```{r libraries, eval=T}
#library
library(Seurat)
library(dplyr)
library(Matrix)
library(cowplot)
library(stringr)
library(SeuratWrappers)
library(readr)
library(PerformanceAnalytics)
library(ggplot2)
```

Let's first load the sequence data


```{r readData1, eval=T}
#CVIDagg6.data <- Read10X(data.dir = "Documents/PhD Thesis Work/Data/Single Cell/CVIDagg6_05Dec2022/agg6_filtered_feature_bc_matrix")
CVIDagg6.data <- Read10X(data.dir = 
	"/home/feng/Windows/windowsD/feng/LAB/MSI/maglione/scRNA_analysis1_20220908/aggr6sub/aggr2212_6sbu/outs/count/filtered_feature_bc_matrix")
CVIDagg6 <- CreateSeuratObject(counts = CVIDagg6.data, project = "CVIDagg6", min.cells = 3, min.features = 200)
ct<-GetAssayData(CVIDagg6, slot="counts")
dat<-GetAssayData(CVIDagg6, slot="data")

#- Add meta.data
mito.genes <- grep(pattern = "^MT-", x = rownames(x = dat), value = TRUE)
percent.mito <- Matrix::colSums(dat[mito.genes, ]) / Matrix::colSums(dat)
CVIDagg6@meta.data$percent.mito <- percent.mito
CVIDagg6@meta.data$sub <- plyr::mapvalues(x = sapply(str_split(rownames(CVIDagg6@meta.data), "[-]"), function(x) x[2]), from = 1:18, to = rep(seq(1,6),c(3,3,3,3,3,3)))
CVIDagg6@meta.data$cvid <- plyr::mapvalues(x = sapply(str_split(rownames(CVIDagg6@meta.data), "[-]"), function(x) x[2]), from = 1:18, to = rep(c("HC","CVID","CVID","HC","CVID","HC"),c(3,3,3,3,3,3)))
CVIDagg6@meta.data$cells <- plyr::mapvalues(x = sapply(str_split(rownames(CVIDagg6@meta.data), "[-]"), function(x) x[2]), from = 1:18, to = rep(c("CD38_high","CD38_mid","CD38_low","CD38_high","CD38_mid","CD38_low","CD38_high","CD38_mid","CD38_low","CD38_high","CD38_low","CD38_mid","CD38_high","CD38_low","CD38_mid","CD38_high","CD38_low","CD38_mid"),c(1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1,1)))

head(CVIDagg6@meta.data) 
```

Now do some visualization to see correlations between 
total number of sequences, unique sequences and total 
cell numbers. 

```{r correlation, eval=T}
stat.cell<-CVIDagg6@meta.data %>% 
	group_by(sub,cvid) %>%
	summarize(cell_counts=n())
stat.seq<-CVIDagg6@meta.data %>% 
	group_by(sub,cvid) %>%
	summarize(total_counts=sum(nFeature_RNA), 
		unique_counts=sum(nCount_RNA))

stat<-stat.cell %>% left_join(stat.seq,by=c("sub","cvid"))
chart.Correlation(stat[,c("cell_counts", 
	"total_counts","unique_counts")])
```

It is no surprising that total number of sequences and unique
sequences are highly correlated. They are also quite well
correlated with the total number of cells. 

To me, it makes more sense to use total number of cells to
normalize the mutation counts. 

Next, read in the mutation data
```{r readData2, eval=T} 
#get number of cells and sequences for each subject
muts<-read_csv("/home/feng/Windows/windowsD/feng/LAB/MSI/maglione/SComatic_output/counts_mutations.csv")
stat <-stat %>% mutate(sub=as.integer(sub))
muts.stat<-stat %>% left_join(muts,by=c("sub"))


```

# 3 analysis and visualization.

We first normalize the counts data using total number of cells.

```{r analysis, eval=T}
muts.stat <- muts.stat %>% 
	mutate(TACI_muts_normed= TACI_muts/cell_counts*1000,
		total_C17_normed=Total_muts_passed_C17/cell_counts*1000,
		total_C22_normed=Total_muts_passed_C22/cell_counts*1000
		)
```

It is hard here to do statistical analysis because the 
sample size is very small, n=3. We can only run non-parametric
Wilcoxon test. 

```{r stats, eval=T}
wilcox.test(TACI_muts_normed~cvid,data=muts.stat )

wilcox.test(total_C17_normed~cvid,data=muts.stat )


wilcox.test(total_C17_normed~cvid,data=muts.stat )
```

Of course, with this small sample size, you can not expect
the Wilcoxon test to give a significant result at the 0.05
level.

But the visualization show that the two populations are 
different. 

- TACI locus
```{r viz, eval=T}
ggplot(data=muts.stat, aes(y=TACI_muts_normed,x=sub,colour=as.factor(cvid)))+
	geom_boxplot() + xlab("Subjects")+ylab("Frequency/1K cells")+
	theme(legend.position="none")
```

- Chromosome 17

```{r viz17, eval=T}
ggplot(data=muts.stat, aes(y=total_C17_normed,x=sub,colour=as.factor(cvid)))+
	geom_boxplot()+ xlab("Subjects")+ylab("Frequency/1K cells")+
	theme(legend.position="none")
```
- Chromosome 22

```{r viz22, eval=T}
ggplot(data=muts.stat, aes(y=total_C22_normed,x=sub,colour=as.factor(cvid)))+
	geom_boxplot()+ xlab("Subjects")+ylab("Frequency/1K cells")+
	theme(legend.position="none")

```

Done!!!

```{r}
sessionInfo()
```